Wrattler: \textnormal{A platform for AI-assisted data science}

### 2018-02-28 report/paper.tex
Data store format section

\subsection{Semantic annotations}
As noted in Section~\ref{sec:wrattler-ds}, annotations can be attached to individual cells, 
entire columns and rows, but also combinations of multiple columns or rows. As an example,
consider the following small table that records data on broadband speed (and illustrates a number
of real-world problems):

\vspace{1em}
\noindent
\begin{tabular}{llc}
\toprule
\textbf{ISP}\qquad\qquad\qquad & \textbf{Postcode}\qquad\qquad & \textbf{Download} \\
\midrule
Virgin & CB43BE & 46.34 \\
BT & CB43PT & N/A \\
Virgin & NW53HL & 41.2 \\
BT & Tottenham & 16.4 \\
\bottomrule
\end{tabular}
\vspace{1em}

\noindent
If we run an AI asistant that performs probabilistic type inference (ptype) on this data set,
it will annotate the table with a number of information:
\begin{itemize}
\item[--] aa
\end{itemize}

{\color{red}
Data store provides a way of annotating cells, columns, combinations of columns and rows.
For example, a column may be annotated with a type that indicates that its values are a mix of
postcodes and city names. A cell can be annotated with an information saying that this value is 
more likely a postcode than a city name. We also need to support annotaitons on a combination of
columns, because three columns can jointly represent an address. A row can be annotated with a 
provenance -- for example, when merging rows from two data sets with different sources.
}

\subsection{Annotation format}
{\color{red}
We want a lightweight annotation format so that implementing an AI assistant or a language
runtime that needs to communicate with the data store is not too hard.
Our annotation format will follow something like \url{http://schema.org}, which defines
a hierarchy of entity types and has a simple way for adding annotations to things.
We will need to define our own hierarchy for schemas for data science.
}

\subsection{Schemas for data science}
{\color{red}
This hierarchy will include some basic things that the data store understands and can 
convert between (integers, floats, unbounded size integers, ISO dates, etc.)

>>>>

\subsection{Semantic annotations}
As noted in Section~\ref{sec:wrattler-ds}, annotations can be attached to individual cells, 
entire columns and rows, but also combinations of multiple columns or rows. As an example,
consider the following small table obtained from a government web site that records data 
on broadband speed (and illustrates a number of real-world problems):

\vspace{1em}
\begin{tabular}{llc}
\toprule
\textbf{ISP}\qquad\qquad\qquad & \textbf{Postcode}\qquad\qquad & \textbf{Download} \\
\midrule
Virgin & CB43BE & 46.34 \\
BT & CB43PT & N/A \\
Virgin & NW53HL & 41.2 \\
BT & London & 16.4 \\
\bottomrule
\end{tabular}
\vspace{1em}

\noindent
As a first step, the data analyst might use the probabilistic type inference (ptype) assistant
in order to infer the types of data in the data set. The assistant needs to be able to annotate
the table with the following information:
%
\begin{itemize}
\item[--] With high certainty, the ISP column is a categorical column with values Virgin and BT;
  with lower certainty, it could also be a string.
\item[--] The column Postcode is most likely a postcode; with lower probability it is a city name
  and with similarly low probability, it is a plain string.
\item[--] The column Download is most likely a number or a missing value represented as N/A.  
\item[--] The N/A cell is annotated as being most likely a missing value, or less likely a plain 
  string. Other cells of the Download column are annotated as most likely being decimal numbers.
\item[--] The cells of the Postcode column are annotated similarly. The first three are most
  likely postcodes, while the last one is most likely a city name, but all also have a non-zero
  probability of being an arbitrary string.
\end{itemize}

\noindent
Assuming the data frame is extracted from a government web page, each row of the data set
could further be annotated with the URL of the data source. This makes it possible to track
provenance at a fine-grained level. If we then append two such data frames, we will know
which row comes from which government data set.

It is worth noting that the data store does not need to fully understand the semantics of such 
annotations. The probabilistic type inference assistant simply adds them and another assistant
can then clean the data and, for example, drop values that do not conform to the most 
probable type. We return to this topic in Section~\ref{sec:datastore-schema}, after discussing
the format of such annotations.

\subsection{Annotation format}
The format of annotations needs to be light-weight so that it can be easily consumed and produced
by AI assistants created in a variety of different programming languages. For this reason, it is
desirable to use a standard file format such as JSON or XML. Furthermore, the format needs to 
be extensible -- although Wrattler should provide a schema for basic annotations, new AI assistants
should be able to freely define and use new kinds of annotations.

An annotation format that satisfies the above constraints can take inspiration from the 
Schema.org project (\url{http://schema.org}), which provides a way of annotating web content
with semantic information. The focus of Schema.org is on higher-level entities (such as 
a person, an event or a place), rather than lower-level information, but the format itself could
be reused by Wrattler.

As an example, consider the following Schema.org description of a place (the following shows a
JSON encoding, but it is worth noting that Schema.org also supports RDF encoding):
%
\begin{equation*}
\begin{array}{l}
\{~~\; \str{@context}: \str{http://schema.org},\\
\quad  \str{@type}: \str{Place},\\
\quad  \str{name}: \str{The Black Lodge},\\
\quad  \str{address}: \{\\
\qquad    \str{@type}: \str{PostalAddress},\\
\qquad    \str{addressLocality}: \str{London},\qquad\qquad\qquad\qquad\qquad~\\
\qquad    \str{postalCode}: \str{NW53HL} ~\}~\}\\
\end{array}
\end{equation*}

\noindent
The \str{@context} attribute defines a namespace of a schema that defines the available
entities. The default namespace, used here, includes two entities that we use in this 
example, namely Place and PostalAddress.

The namespace mechanism provides a convenient way of extending annotations -- Wrattler can
define a core schema for common information that it needs, but new AI assistants can 
extend it via custom namespaces.

The Schema.org format also makes it easy to attach additional information to an existing
value. In the example above, the original value might have been ``The Black Lodge'', but 
an AI assistant -- possibly using entity linkage -- inferred and extended it with postal address.

\subsection{Schemas for data science}
\label{sec:datastore-schema}

{\color{red}
This hierarchy will include some basic things that the data store understands and can 
convert between (integers, floats, unbounded size integers, ISO dates, etc.)
